---
title: "My Tran - Research Project Sp2023"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2023-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Preparation:**

```{r, echo=T, message = F, warning = F}
library(tidyverse, quietly = T)
library('scales')
library(ggplot2)
library(plotly)
```

# **Data:**

```{r, warning = F}
library(readxl)
data <- read_excel("Arthropod Master Report 2015_2020.xlsx",na=c("","NA"))
data <- data.frame(data)
attach(data)
dim(data)
```

# **Get familiar with data:**

## **First subset of data**

```{r}
library(tidyverse)
library(lubridate)
data <- data %>% mutate(Date = make_date(year, month, day))
```

```{r}
data1 <- data %>%
  select(DateTime, SpeciesAbbr1, 
         TotFemCount1, TotMaleCount1, TrapType) %>%
  na.omit()
  
data1$TotFemCount1 <- as.numeric(data1$TotFemCount1)
data1$TotMaleCount1 <- as.numeric(data1$TotMaleCount1)
sapply(data1, class)
```

### *Analyze the quality and quantity of species.*

```{r}
total_female = sum(na.omit(data1$TotFemCount1)) #amount of female species

total_male = sum(na.omit(data1$TotMaleCount1)) #amount of female species

total_species = total_male + total_female #amount of species which were collected

rate_female = percent(total_female / total_species) #rate of female species 

rate_male = percent(total_male/total_species) #rate of male species
```

The amount of female species is `r toString(total_female)`

The amount of male species is: `r toString(total_male)`

The total amount of species which were collected is: `r toString(total_species)`

The rate of female species is: `r toString(rate_female)` 

The rate of male species is: `r toString(rate_male)`

```{r}
types = data1 %>% group_by(SpeciesAbbr1) %>% summarise() %>% nrow() #amount of type
```

The total amount of types of species which were collected is: `r toString(types)`

```{r}
data1 %>% summarise(total_female = sum(na.omit(TotFemCount1)),
                    total_male = sum(na.omit(TotMaleCount1)))
```

```{r echo=FALSE, fig.height=12, fig.width=12, message=FALSE}
library(ggplot2)
library(reshape)
data1_2 <- data1 %>% select(SpeciesAbbr1, TotFemCount1, TotMaleCount1) %>% 
  group_by(SpeciesAbbr1) %>%
  summarise(female = sum(TotFemCount1), male = sum(TotMaleCount1)) %>%
  arrange(desc(female)) %>%
  head(10)
data1_2 <- data.frame(data1_2)
df2 <- reshape::melt(data1_2, id = c("SpeciesAbbr1"))
df2 %>% ggplot(aes(x = SpeciesAbbr1, y = value, fill = variable)) + 
  geom_bar(stat = "identity") + facet_wrap(~ variable) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



```

### *Season*

```{r, echo=F,fig.height=10, fig.width=12, message=F}
data1 %>%
  plot_ly(x = ~month(data1$Date)) %>% 
  add_histogram(color = I("pink"), 
                stroke = I("black"), 
                span = I(2), 
                alpha = 0.6) %>% 
  group_by(month(data1$Date)) %>%
  summarise(Species_Count = n()) %>% 
  layout(title = "Histogram of Collected Species by Month", xaxis = list(title = "Month"), yaxis = list(title = "Amount of Species")) 
```

```{r}
df1 <- data1 %>% group_by(weekday = wday(data1$DateTime), TrapType) %>% 
  summarise (total = sum(TotMaleCount1)+sum(TotFemCount1)) 
df1 %>%
  ggplot(aes(x= weekday,y=total,group=TrapType,color=TrapType)) + 
  geom_point(size = 3) + 
  geom_line(size = 0.5) + 
  facet_wrap(~TrapType,nrow=1) + 
  theme(legend.position="none") +
  scale_x_discrete(breaks=c(1,2,3,4,5,6,7))


```

```{r}
library(ggplot2)
theme_set(theme_minimal())
library(tidyverse)
df2 <- data1 %>% group_by(year = year(data1$DateTime)) %>% 
  summarise (female = sum(TotFemCount1), 
             male = sum(TotMaleCount1),
             total = sum(TotMaleCount1)+sum(TotFemCount1)) 
df2 %>%
  ggplot(aes(x=year)) + 
  geom_line(aes(y=female),color='darkred') +
  geom_line(aes(y = male), color="steelblue") +
  labs(y='Species Amount',
       x='Year',
       title='Collected Species by Year') 


```

```{r}
library(plotly)
library(dplyr)
data1 %>% 
  group_by(Year = year(data1$DateTime)) %>% 
  summarise(Species_Amount = n()) %>% 
  ggplot(aes(Year,Species_Amount)) + geom_point() + geom_smooth()
```

```{r}
#How many positive happens?
data3 <- data %>% group_by(WNVClass,SLEClass,CHKClass,DENClass,ZKAClass) %>%
  summarise(total_mosquitoes = sum(sum(TotFemCount1),sum(TotMaleCount1)))
```

```{r}
#data %>% group_by(WNVClass) %>% count(month_year) %>% arrange(month_year)
#map for month, n (no Negative)
```

```{r}
#How many positive over year?
data %>% group_by(Year) %>% 
  filter(WNVClass == 'Positive') %>%
  summarise(total_mosquitoes = sum(sum(TotFemCount1),sum(TotMaleCount1))) 
```



```{r}
data2 <- data %>% filter(Zip!=7737) %>% group_by(Zip) %>% na.omit() %>%
  mutate(total_mosquitoes = sum(sum(TotFemCount1),sum(TotMaleCount1))) 
```
```{r}
library(sf)
zipcode <- st_read('cb_2020_us_zcta520_500k.shx',layer = "cb_2020_us_zcta520_500k")
```


```{r}
library(tigris)


houston_zipcodes <- tigris:: zctas(cb=F,class='sf',starts_with = 77)

data2$Zip <- as.character(data2$Zip) 


heat_map_of_Houston <- houston_zipcodes %>% left_join(data2, by = c("ZCTA5CE20" = 'Zip')) %>% na.omit()
```

```{r,fig.width=19, fig.height=16}

ggplot(heat_map_of_Houston) + geom_sf(aes(fill = total_mosquitoes)) + scale_fill_viridis_c(option = "inferno" , begin = 0.1, label = scales::comma) + stat_sf_coordinates() + theme(legend.position="right",
        plot.title = element_text(hjust = 0.5,color = "dark red", size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, color = "dark red", size = 14, face = "bold"),
        plot.caption = element_text(color = "Gray60"))+ 
  guides(fill = guide_legend(title = "Mosquitoes Count", title.position = "top", 
                             title.theme =element_text(size = 10, face = "bold",
                                                       colour = "Black",angle = 0)))  + 
  geom_sf_text(aes(label = ZCTA5CE20), colour = "grey", alpha = 0.6) + 
  labs(title = "Mosquitoes Count by Zipcode in Houston", subtitle = "2015 to 2020")
```


```{r}
#Weather
##Weather Data from The National Oceanic and Atmospheric Administration (NOAA)'s Climate Data Online
install.packages("xts")
install.packages("tsbox")
install.packages("forecast")

```

```{r}
weather <- read.csv('weather.csv')
names(weather)
range(weather$DATE)
```

```{r}
data4 <-weather %>% group_by(DATE) %>% filter (NAME == 'HOUSTON NATIONAL WEATHER SERVICE OFFICE, TX US') %>% select(DATE, NAME, PRCP, LATITUDE, LONGITUDE) 
```


```{r}
data4$Date <- as.Date(data4$DATE)
# Merge the data sets based on date-time column
merged_data <- data %>% inner_join(data4, by=c('Date'))
merged_data <- merged_data %>% select(SpeciesAbbr1, TotFemCount1, TotMaleCount1, Date, DateTime, PRCP) %>% 
  group_by(Date, PRCP) %>%
  summarise(total = sum(TotFemCount1) + sum(TotMaleCount1)) %>% 
  na.omit(total)
```

```{r}
model <- lm(total ~ PRCP, merged_data)
summary(model)
```

